Component response rate variation drives stability in large complex systems

### 2018-07-05 notebook/ms_for_review.Rmd
Adjustments for Science

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie    ( alexander.duthie@stir.ac.uk )"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA
output:
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - science.csl
    reference_docx: docx_template.docx
  pdf_document:
    fig_caption: yes
    keep_tex: no
  html_document: default
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
link-citations: no
linkcolor: blue
csl: nature.csl
biblio-style: apalike
---

@@

@@

In 1972, May [@May1972] first demonstrated that randomly assembled systems of sufficient complexity are almost inevitably unstable given infinitesimally small perturbations. Complexity in this case is defined by the size of the system (i.e., the number of potentially interacting components; $S$), its connectance (i.e., the probability that one component will interact with another; $C$), and the variance of interaction strengths ($\sigma^{2}$)[@Allesina2012]. May's finding that the probability of local stability falls to near zero given a sufficiently high threshold of $\sigma\sqrt{SC}$ has profound consequences across multiple disciplines, raising the question of how complex systems in, e.g., ecology[@Mougi2012; @Allesina2012; @Allesina2015; @Grilli2017] or banking [@May2008; @Haldane2011; @Bardoscia2017] are predicted to persist or change.

Randomly assembled complex systems can be represented as large square matrices ($M$) with $S$ components (e.g., species[@Allesina2012] or banks[@Haldane2011]). One element of such a matrix, $M_{ij}$, defines how component $j$ affects component $i$ in the system at a point of equilibrium[@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation (e.g., carrying capacity in ecological communities). Traditionally, off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma^{2}$; diagonal elements are set to -1[@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis, with the system being stable if the real parts of all eigenvalues ($\lambda$) of $M$ are negative ($\max\left(\Re(\lambda)\right) < 0$)[@May1972; @Allesina2012]. In a large system (high $S$), eigenvalues are distributed uniformly[@Tao2010] within a circle centred at $\Re = -1$ (the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma\sqrt{SC}$[@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S$, $C$, and $\sigma^{2}$ increase.

May's[@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that individual components respond to perturbations of the system at the same rate ($\gamma$), but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of individuals, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show why, despite higher $\sigma$, complex systems in which $\gamma$ varies are actually more likely to be stable, especially when $S$ is high.

@@

```{r, echo = FALSE}
A_comp <- NULL;
A_dat  <- rnorm(n = 1000000, mean = 0, sd = 0.4);
A_mat  <- matrix(data = A_dat, nrow = 1000);
C_dat  <- rbinom(n = 1000 * 1000, size = 1, prob = 1);
C_mat  <- matrix(data = C_dat, nrow = 1000, ncol = 1000);
A_mat     <- A_mat * C_mat;
gammas <- runif(n = 1000, min = 0, max = 2);
mu_gam <- mean(gammas);
diag(A_mat) <- -1;
A1     <- gammas * A_mat;
A0     <- mu_gam * A_mat;
A0_e   <- eigen(A0)$values;
A0_r   <- Re(A0_e);
A0_i   <- Im(A0_e);
A1_e   <- eigen(A1)$values;
A1_r   <- Re(A1_e);
A1_i   <- Im(A1_e);

A0_vm       <- A0;
diag(A0_vm) <- NA;
A0vec       <- as.vector(A0_vm);
A0vec       <- A0vec[is.na(A0vec) == FALSE];
A1_vm       <- A1;
diag(A1_vm) <- NA;
A1vec       <- as.vector(A1_vm);
A1vec       <- A1vec[is.na(A1vec) == FALSE];
```

Higher stability in systems with variation in $\gamma$ can be observed by sampling $\gamma_{i}$ values from various distributions. I now focus on a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supplementary Information for other distributions, which give similar results). As with the case of $\gamma = \{1.95, 0.5\}$ (Fig. 1b), $E[\gamma] = 1$ when $\gamma \sim \mathcal{U}(0, 2)$, allowing comparison of $M$ before and after variation in component response rate. Figure 2 shows a comparison of eigenvalue distributions given $S = 1000$, $C = 1$, and $\sigma = 0.4$. As expected [@Tao2010], when $\gamma = 1$, eigenvalues are distributed uniformly in a circle centred at ($-1, 0$) with a radius of $\sigma\sqrt{SC} =$ `r round(sigma*sqrt(S*C), digits = 3)`. Uniform variation in $\gamma$ leads to a non-uniform distribution of eigenvalues, some of which are clustered tightly around the centre of the distribution, but others of which are spread outside the former radius of `r round(sigma*sqrt(S*C), digits = 3)` (red circle Fig 2b). This larger radius occurs because the addition of $Var(\gamma)$ increases the realised $\sigma$ of $M$. The clustering and spreading of eigenvalues introduced by $Var(\gamma)$ can destabilise previously stable systems or stabilise systems that are otherwise unstable. But where systems are otherwise too complex to be stable given $\gamma = 1$, the effect of $Var(\gamma)$ can often lead to stability above May's[@May1972; @Allesina2012] threshold $\sigma\sqrt{SC} < 1$.

@@

To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a genetic algorithm (the space of possible $\gamma$ values was too large to search exhaustively[@Hamblin2013]; see Supplementary Information). For each of 40000 random $M$ ($\sigma = 0.4$, $C = 1$), the genetic algorithm initialised 1000 different sets of $\gamma \sim \mathcal{U}(0, 2)$ values of size $S$. Eigenanalysis was performed on $M$ using each set of $\gamma$ values, and the 20 sets with the lowest $\max\left(\Re(\lambda)\right)$ each produced 50 clonal offspring with subsequent mutation and crossover between the resulting new population of 1000 $\gamma$ sets. The genetic algorithm terminated if a stable $M$ was found, 20 generations occurred, or a convergence criteria of minimum fitness increase between generations was satisfied. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were up to four orders of magnitude higher than when $\gamma = 1$ (see Supplementary Information), meaning that stability could often be achieved by manipulating $S$ $\gamma$ values rather than $S \times S$ $M$ elements. Hence, managing the response rates of system components in a targetted way can potentially facilitate the stabilisation of complex systems through a reduction in dimensionality.

I have focused broadly on random complex systems, but it is also worthwhile to consider more restricted interactions such as those of specific ecological networks[@Allesina2012]. These include systems in which all interactions are negative (competitive networks), positive (mutualist networks), or $i$ and $j$ pairs have opposing signs (predator-prey networks). In general, competitive and mutualist networks tend to be destabilising, and predator-prey network tend to be stabilising[@Allesina2011]. When $Var(\gamma)$ is applied to each, the proportion of stable competitive and predator-prey networks increases, but the proportion of stable mutualist networks does not (see Supplementary Information). Additionally, when each component of $M$ is interpreted as a unique species and given a random intrinsic growth rate[@Dougoud2018], feasibility is not increased by $Var(\gamma)$, suggesting that variation in species generation time might be unlikely to drive stability in purely multi-species networks (see Supplementary Information).

My results show that complex systems are more likely to be stable when the response rates of system components vary. These results are broadly applicable to complex biological and social networks.

@@

**Figure 1: Example distribution of eigenvalues before (a) and after (b) separating a randomly generated complex system into fast ($\boldsymbol{\gamma} = 1.95$) and slow ($\boldsymbol{\gamma} = 0.05$) component response rates.** Each panel shows the same system where $S = 200$, $C = 0.05$, and $\sigma = 0.4$, and in each case $E[\gamma] = 1$ (i.e., only the distribution of $\gamma$ differs between panels).  **a.** Eigenvalues plotted when all $\gamma = 1$; distributions of points are uniformly distributed within the grey circle with a radius of $\sigma\sqrt{SC} =$ `r round(sqrt(200) * sd(A0vec), digits =3)` centred at -1 on the real axis. **b.** Eigenvalues plotted when half $\gamma = 1.95$ and half $\gamma = 0.05$; distributions of points can be partitioned into one large circle of $\sigma\sqrt{SC} =$ `r round(sqrt(100) * sd(A1vec[fhalf]), digits = 3)` centred at $\gamma = -1.95$ and one small circle of $\sigma\sqrt{SC} =$ `r round(sqrt(100) * sd(A1vec[shalf]), digits = 3)` centred at $\gamma = -0.05$. In a, the maximum real eigenvalue $\max\left(\Re(\lambda)\right) =$ `r format(max(A0_r), scientific = FALSE)`, while in b $\max\left(\Re(\lambda)\right) =$ `r format(max(A1_r), scientific = FALSE)`, meaning that the complex system in b but not a is stable because in b $\max\left(\Re(\lambda)\right) < 0$. In 1 million randomly generated complex systems under the same parameter values, `r sum(pr_st[,1])` was stable when $\gamma = 1$ while `r sum(pr_st[,2])` were stable when $\gamma = \{1.95, 0.05\}$. Overall, complex systems that are separated into fast versus slow components tend to be more stable than otherwise identical systems with identical component response rates.

@@

```{r, echo = FALSE}
A_comp <- NULL;
A_dat  <- rnorm(n = 1000000, mean = 0, sd = 0.4);
A_mat  <- matrix(data = A_dat, nrow = 1000);
C_dat  <- rbinom(n = 1000 * 1000, size = 1, prob = 1);
C_mat  <- matrix(data = C_dat, nrow = 1000, ncol = 1000);
A_mat     <- A_mat * C_mat;
gammas <- runif(n = 1000, min = 0, max = 2);
mu_gam <- mean(gammas);
diag(A_mat) <- -1;
A1     <- gammas * A_mat;
A0     <- mu_gam * A_mat;
A0_e   <- eigen(A0)$values;
A0_r   <- Re(A0_e);
A0_i   <- Im(A0_e);
A1_e   <- eigen(A1)$values;
A1_r   <- Re(A1_e);
A1_i   <- Im(A1_e);

A0_vm       <- A0;
diag(A0_vm) <- NA;
A0vec       <- as.vector(A0_vm);
A0vec       <- A0vec[is.na(A0vec) == FALSE];
A1_vm       <- A1;
diag(A1_vm) <- NA;
A1vec       <- as.vector(A1_vm);
A1vec       <- A1vec[is.na(A1vec) == FALSE];
```

**Figure 2: Distributions of eigenvalues before (a) and after (b) introducing variation in component response rate ($\boldsymbol{\gamma}$) in complex systems.** Each panel show the same system where $S = 1000$, $C = 1$, and $\sigma = 0.4$. **a.** Eigenvalues plotted in the absence of $Var(\gamma)$ where $E[\gamma] = 1$, versus **b.** eigenvalues plotted given $\gamma \sim \mathcal{U}(0, 2)$, which increases the variance of interaction strengths ($\sigma^{2}$) but clusters eigenvalues toward the distribution's centre (-1, 0). Black and red elipses in both panels show the circle centred on the distribution in panels a and b, respectively, which have a radius of $\sigma \sqrt{SC}$. Proportions of $\Re(\lambda) < 0$ are `r sum(A0_r < 0) / length(A0_r)` and `r sum(A1_r < 0) / length(A1_r)` for a and b, respectively.

```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 9}
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5), oma = c(5, 5, 0, 0));
plot(A0_r, A0_i, xlim = c(-16.5, 15.5), ylim = c(-16.5,15.5), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1);
vl <- seq(from = 0, to = 2*pi, by = 0.001);
x0 <- sqrt(1000) * sd(A0vec) * cos(vl) + mean(diag(A0));
y0 <- sqrt(1000) * sd(A0vec) * sin(vl);
x1 <- sqrt(1000) * sd(A1vec) * cos(vl) + mean(diag(A1));
y1 <- sqrt(1000) * sd(A1vec) * sin(vl);
text(x = -15.5, y = 19, labels = "a", cex = 2);
points(x = x0, y = y0, type = "l", lwd = 3);
points(x = x1, y = y1, type = "l", col = "red", lwd = 3, lty = "dashed");
plot(A1_r, A1_i, xlim = c(-16.5, 15.5), ylim = c(-16.5,15.5), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1, col = "red",
     yaxt = "n");
text(x = -15.5, y = 19, labels = "b", cex = 2);
points(x = x1, y = y1, type = "l", col = "red", lwd = 3)
points(x = x0, y = y0, type = "l", lwd = 3, lty = "dashed");
mtext(side = 1, "Real", outer = TRUE, line = 3, cex = 2);
mtext(side = 2, "Imaginary", outer = TRUE, line = 2.5, cex = 2);
```

>>>>

---
title: "Component response rate variation drives stability in large complex systems"
author: "A. Bradley Duthie    ( alexander.duthie@stir.ac.uk )"
date: Biological and Environmental Sciences, University of Stirling, Stirling, UK,
  FK9 4LA
output:
  pdf_document:
    fig_caption: yes
    keep_tex: no
  html_document: default
  word_document:
    fig_caption: yes
    pandoc_args:
    - --csl
    - science.csl
    reference_docx: docx_template.docx
bibliography: references.bib
header-includes:
- \usepackage{amsmath}
- \usepackage{natbib}
- \usepackage{lineno}
- \usepackage[utf8]{inputenc}
- \linenumbers
- \bibliographystyle{amnatnat}
linestretch: 2
link-citations: no
linkcolor: blue
csl: science.csl
biblio-style: apalike
---

@@

**The stability of a complex system generally decreases with increasing system size and interconnectivity, a counter-intuitive result of widespread importance in the physical, life, and social sciences. Nevertheless, stabilising mechanisms in complex systems remain under-developed, and the effect of variation in the response rate of individual system components remains an open problem. Here I vary the component response rates ($\boldsymbol{\gamma}$) of randomly generated complex systems. I show that when component response rates vary, the potential for system stability is markedly increased. Variation in $\boldsymbol{\gamma}$ becomes increasingly important as system size increases, such that the largest stable complex systems would be unstable if not for $\boldsymbol{Var(\gamma)}$. My results reveal a previously unconsidered driver of system stability that is likely to be pervasive across all complex systems.**

<!---

@@

--->

In 1972, May [@May1972] first demonstrated that randomly assembled systems of sufficient complexity are almost inevitably unstable given infinitesimally small perturbations. Complexity in this case is defined by the size of the system (i.e., the number of potentially interacting components; $S$), its connectance (i.e., the probability that one component will interact with another; $C$), and the variance of interaction strengths ($\sigma^{2}$) [@Allesina2012]. May's finding that the probability of local stability falls to near zero given a sufficiently high threshold of $\sigma\sqrt{SC}$ is broadly relevant for understanding the dynamics and persistence of systems such as ecological [@May1972; @Allesina2012; @Mougi2012; @Allesina2015; @Grilli2017], neurological [@Gray2008; @Gray2009], biochemical [@Rosenfeld2009; @MacArthur2010], and socio-economic [e.g., banking; @May2008; @Haldane2011; @Suweis2014; @Bardoscia2017] networks. As such, identifying general principles that drive stability in complex systems is of wide-ranging importance.

Randomly assembled complex systems can be represented as large square matrices ($M$) with $S$ components (e.g., species[@Allesina2012] or banks[@Haldane2011]). One element of such a matrix, $M_{ij}$, defines how component $j$ affects component $i$ in the system at a point of equilibrium [@Allesina2012]. Off-diagonal elements ($i \neq j$) therefore define interactions between components, while diagonal elements ($i = j$) define component self-regulation (e.g., carrying capacity in ecological communities). Traditionally, off-diagonal elements are assigned non-zero values with a probability $C$, which are sampled from a distribution with variance $\sigma^{2}$; diagonal elements are set to -1[@May1972; @Allesina2012; @Allesina2015]. Local system stability is assessed using eigenanalysis, with the system being stable if the real parts of all eigenvalues ($\lambda$) of $M$ are negative ($\max\left(\Re(\lambda)\right) < 0$) [@May1972; @Allesina2012]. In a large system (high $S$), eigenvalues are distributed uniformly [@Tao2010] within a circle centred at $\Re = -1$ (the mean value of diagonal elements) and $\Im = 0$, with a radius of $\sigma\sqrt{SC}$ [@May1972; @Allesina2012; @Allesina2015] (Figs 1a and 2a). Local stability of randomly assembled systems therefore becomes increasingly unlikely as $S$, $C$, and $\sigma^{2}$ increase.

May's [@May1972; @Allesina2012] stability criterion $\sigma\sqrt{SC} < 1$ assumes that individual components respond to perturbations of the system at the same rate ($\gamma$), but this is highly unlikely in any complex system. In ecological communities, for example, the rate at which population density changes following perturbation will depend on the generation time of individuals, which might vary by orders of magnitude among species. Species with short generation times will respond quickly (high $\gamma$) to perturbations relative to species with long generation times (low $\gamma$). Similarly, the speed at which individual banks respond to perturbations in financial networks, or individuals or institutions respond to perturbations in complex social networks, is likely to vary. The effect of such variance has not been investigated in complex systems theory. Intuitively, variation in $\gamma$ might be expected to decrease system stability by introducing a new source of variation into the system and thereby increasing $\sigma$. Here I show why, despite higher $\sigma$, complex systems in which $\gamma$ varies are actually more likely to be stable, especially when $S$ is high.

@@

```{r, echo = FALSE}
A_comp <- NULL;
A_dat  <- rnorm(n = 1000000, mean = 0, sd = 0.4);
A_mat  <- matrix(data = A_dat, nrow = 1000);
C_dat  <- rbinom(n = 1000 * 1000, size = 1, prob = 0.05);
C_mat  <- matrix(data = C_dat, nrow = 1000, ncol = 1000);
A_mat     <- A_mat * C_mat;
A_mat  <- species_interactions(A_mat, type = 0);
gammas <- c(rep(0.05, 500), rep(1.95, 500));

mu_gam <- mean(gammas);
diag(A_mat) <- -1;
A1     <- gammas * A_mat;
A0     <- mu_gam * A_mat;
A0_e   <- eigen(A0)$values;
A0_r   <- Re(A0_e);
A0_i   <- Im(A0_e);
A1_e   <- eigen(A1)$values;
A1_r   <- Re(A1_e);
A1_i   <- Im(A1_e);

A0_vm       <- A0;
diag(A0_vm) <- NA;
A0vec       <- as.vector(A0_vm);
A0vec       <- A0vec[is.na(A0vec) == FALSE];
A1_vm       <- A1;
diag(A1_vm) <- NA;
A1vec       <- as.vector(A1_vm[1:500,]);
A1vec       <- A1vec[is.na(A1vec) == FALSE];
```

Higher stability in systems with variation in $\gamma$ can be observed by sampling $\gamma_{i}$ values from various distributions. I now focus on a uniform distribution where $\gamma \sim \mathcal{U}(0, 2)$ (see Supplementary Information for other distributions, which give similar results). As with the case of $\gamma = \{1.95, 0.5\}$ (Fig. 1b), $E[\gamma] = 1$ when $\gamma \sim \mathcal{U}(0, 2)$, allowing comparison of $M$ before and after variation in component response rate. Figure 2 shows a comparison of eigenvalue distributions given $S = 1000$, $C = 1$, and $\sigma = 0.4$. As expected [@Tao2010], when $\gamma = 1$, eigenvalues are distributed uniformly in a circle centred at ($-1, 0$) with a radius of $\sigma\sqrt{SC} =$ `r round(sigma*sqrt(S*C), digits = 3)`. Uniform variation in $\gamma$ leads to a non-uniform distribution of eigenvalues, some of which are clustered tightly around the centre of the distribution, but others of which are spread outside the former radius of `r round(sigma*sqrt(S*C), digits = 3)` (red circle Fig 2b). This larger radius occurs because the addition of $Var(\gamma)$ increases the realised $\sigma$ of $M$. The clustering and spreading of eigenvalues introduced by $Var(\gamma)$ can destabilise previously stable systems or stabilise systems that are otherwise unstable. But where systems are otherwise too complex to be stable given $\gamma = 1$, the effect of $Var(\gamma)$ can often lead to stability above May's [@May1972; @Allesina2012] threshold $\sigma\sqrt{SC} < 1$.

@@

To further investigate the potential of $Var(\gamma)$ to be stabilising, I used a genetic algorithm (the space of possible $\gamma$ values was too large to search exhaustively [@Hamblin2013]; see Supplementary Information). For each of 40000 random $M$ ($\sigma = 0.4$, $C = 1$), the genetic algorithm initialised 1000 different sets of $\gamma \sim \mathcal{U}(0, 2)$ values of size $S$. Eigenanalysis was performed on $M$ using each set of $\gamma$ values, and the 20 sets with the lowest $\max\left(\Re(\lambda)\right)$ each produced 50 clonal offspring with subsequent mutation and crossover between the resulting new population of 1000 $\gamma$ sets. The genetic algorithm terminated if a stable $M$ was found, 20 generations occurred, or a convergence criteria of minimum fitness increase between generations was satisfied. Across $S = \{2, 3, ..., 39, 40\}$, sets of $\gamma$ values were found that resulted in stable systems with probabilities that were up to four orders of magnitude higher than when $\gamma = 1$ (see Supplementary Information), meaning that stability could often be achieved by manipulating $S$ $\gamma$ values rather than $S \times S$ $M$ elements. Hence, managing the response rates of system components in a targetted way can potentially facilitate the stabilisation of complex systems through a reduction in dimensionality.

I have focused broadly on random complex systems, but it is also worthwhile to consider more restricted interactions such as those of specific ecological networks[@Allesina2012]. These include systems in which all interactions are negative (competitive networks), positive (mutualist networks), or $i$ and $j$ pairs have opposing signs (predator-prey networks). In general, competitive and mutualist networks tend to be destabilising, and predator-prey network tend to be stabilising[@Allesina2011]. When $Var(\gamma)$ is applied to each, the proportion of stable competitive and predator-prey networks increases, but the proportion of stable mutualist networks does not (see Supplementary Information). Additionally, when each component of $M$ is interpreted as a unique species and given a random intrinsic growth rate [@Dougoud2018], feasibility is not increased by $Var(\gamma)$, suggesting that variation in species generation time might be unlikely to drive stability in purely multi-species networks (see Supplementary Information).

My results show that complex systems are more likely to be stable when the response rates of system components vary. These results are broadly applicable to understanding stability of complex networks in the physical, life, and social sciences.

@@

**Figure 1: Example distribution of eigenvalues before (a) and after (b) separating a randomly generated complex system into fast ($\boldsymbol{\gamma} = 1.95$) and slow ($\boldsymbol{\gamma} = 0.05$) component response rates.** Each panel shows the same system where $S = 200$, $C = 0.05$, and $\sigma = 0.4$, and in each case $E[\gamma] = 1$ (i.e., only the distribution of $\gamma$ differs between panels).  **a.** Eigenvalues plotted when all $\gamma = 1$; distributions of points are uniformly distributed within the grey circle with a radius of $\sigma\sqrt{SC} =$ `r round(sqrt(200) * sd(A0vec), digits =3)` centred at -1 on the real axis. **b.** Eigenvalues plotted when half $\gamma = 1.95$ and half $\gamma = 0.05$; distributions of points can be partitioned into one large circle centred at $\gamma = -1.95$ and one small circle centred at $\gamma = -0.05$. In a, the maximum real eigenvalue $\max\left(\Re(\lambda)\right) =$ `r format(max(A0_r), scientific = FALSE)`, while in b $\max\left(\Re(\lambda)\right) =$ `r format(max(A1_r), scientific = FALSE)`, meaning that the complex system in b but not a is stable because in b $\max\left(\Re(\lambda)\right) < 0$. In 1 million randomly generated complex systems under the same parameter values, `r sum(pr_st[,1])` was stable when $\gamma = 1$ while `r sum(pr_st[,2])` were stable when $\gamma = \{1.95, 0.05\}$. Overall, complex systems that are separated into fast versus slow components tend to be more stable than otherwise identical systems with identical component response rates.

@@

```{r, echo = FALSE}
A_comp <- NULL;
A_dat  <- rnorm(n = 1000000, mean = 0, sd = 0.4);
A_mat  <- matrix(data = A_dat, nrow = 1000);
C_dat  <- rbinom(n = 1000 * 1000, size = 1, prob = 0.05);
C_mat  <- matrix(data = C_dat, nrow = 1000, ncol = 1000);
A_mat     <- A_mat * C_mat;
A_mat  <- species_interactions(A_mat, type = 0);
gammas <- runif(n = 1000, min = 0, max = 2);

mu_gam <- mean(gammas);
diag(A_mat) <- -1;
A1     <- gammas * A_mat;
A0     <- mu_gam * A_mat;
A0_e   <- eigen(A0)$values;
A0_r   <- Re(A0_e);
A0_i   <- Im(A0_e);
A1_e   <- eigen(A1)$values;
A1_r   <- Re(A1_e);
A1_i   <- Im(A1_e);

A0_vm       <- A0;
diag(A0_vm) <- NA;
A0vec       <- as.vector(A0_vm);
A0vec       <- A0vec[is.na(A0vec) == FALSE];
```

**Figure 2: Distributions of eigenvalues before (a) and after (b) introducing variation in component response rate ($\boldsymbol{\gamma}$) in complex systems.** Each panel show the same system where $S = 1000$, $C = 0.05$, and $\sigma = 0.4$. **a.** Eigenvalues plotted in the absence of $Var(\gamma)$ where $E[\gamma] = 1$, versus **b.** eigenvalues plotted given $\gamma \sim \mathcal{U}(0, 2)$, which increases the variance of interaction strengths ($\sigma^{2}$) but also creates a cluster of eigenvalues toward the distribution's centre (-1, 0). Black elipses in both panels show the circle centred on the distribution in panel a. Proportions of $\Re(\lambda) < 0$ are `r sum(A0_r < 0) / length(A0_r)` and `r sum(A1_r < 0) / length(A1_r)` for a and b, respectively.

```{r, eval = TRUE, echo = FALSE, fig.height = 6, fig.width = 9}
par(mfrow = c(1, 2), mar = c(0.5, 0.5, 0.5, 0.5), oma = c(5, 5, 0, 0));
plot(A0_r, A0_i, xlim = c(-6, 2.25), ylim = c(-5,4.25), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1);
vl <- seq(from = 0, to = 2*pi, by = 0.001);
x0 <- sqrt(1000) * sd(A0vec) * cos(vl) + mean(diag(A0));
y0 <- sqrt(1000) * sd(A0vec) * sin(vl);
text(x = -15.5, y = 19, labels = "a", cex = 2);
points(x = x0, y = y0, type = "l", lwd = 3);
plot(A1_r, A1_i, xlim = c(-6, 2.25), ylim = c(-5,4.25), pch = 4, cex = 0.7,
     xlab = "", ylab = "", cex.lab = 1.5, cex.axis = 1.5, asp = 1, col = "red",
     yaxt = "n");
text(x = -15.5, y = 19, labels = "b", cex = 2);
points(x = x0, y = y0, type = "l", lwd = 3, lty = "solid");
mtext(side = 1, "Real", outer = TRUE, line = 3, cex = 2);
mtext(side = 2, "Imaginary", outer = TRUE, line = 2.5, cex = 2);
```
